Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-21 paper/paper.tex
Fixes

\newmintinline[coq]{coq}{fontsize=\small}
\newmintinline[ocaml]{OCaml}{fontsize=\small}
\newminted{coq}{fontsize=\small}
\newminted{ocaml}{fontsize=\small}

@@

The container encoding \coq{ITreeF} has two shapes, one for leaves and one for
nodes, and nodes have two positions, one for each sublist \coq{l r : X} in
\coq{i_node}.
\begin{coqcode}
Inductive Tshape L A := | Leaf | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
The container for \coq{ITreeF} is denoted by \coq{TreeF}. For convenience, we
define a series of wrappers \coq{a_leaf} and \coq{a_node} to construct these
containers.  Now, the container for quicksort is given by \coq{TreeF int}.
In the definition below, we use \coq{posL} and \coq{posR} as wrappers for
\coq{Lbranch} and \coq{Rbranch} with validity proofs.
\begin{coqcode}
Definition qsplit_f (x : list int) : App (TreeF int) (list int) :=
  match x with
  | nil => a_leaf tt
  | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in a_node h l r
  end.

Definition merge_f (x : App (TreeF int) (list int)) : list int :=
  let (sx, kx) := x in
  match sx return (Container.Pos sx -> _) -> _ with
  | Leaf _ _ => fun _ => nil
  | Node _ h => fun k => List.app (k (posL h)) (h :: k (posR h))
  end kx.
\end{coqcode}
The proofs that these are proper morphisms (called \coq{merge} and \coq{qsplit})
are automatically discharged by our tactics, but we still have to prove that the
coalgebra \coq{qsplit} is recursive. In this specific case we have to prove that
the two sublists are smaller than the input list. In general, we are required to
prove that the input is related to the data inside the structure functor, that
is that the element of the carrier of the coalgebra is related to the content of
the functor by some total relation. Here we give a specialised property, the
full general property can be found in the formalisation:
\begin{coqcode}
Definition respects_relation `{setoid A} (c : Coalg F B) B} (R : B -> B -> Prop)
:= forall x (p : Pos (shape (c x))), R (length (cont (c x) p)) (length x).
\end{coqcode}

The resulting code extracted from \coq{hylo merge qsplit} is similar to a
hand-written implementation:
\begin{ocamlcode}
let rec qsort = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
    let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
    app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
The correctness proof is done by proving the following statement:
\begin{coqcode}
forall (l : list int), Sorted (hylo merge qsplit l) /\ Perm l (hylo merge qsplit l).
\end{coqcode}
that is any list produced by quicksort is sorted and is some permutation of the original one.

@@

\subsubsection{Fusing a divide-and-conquer computation}
We now show how we can use program calculation techniques to fuse a traversal
with a  divide-and-conquer algorithm to obtain a new program that only performs
recursion once instead of twice. In particular, we can map a function to the
result of quicksort and obtain a new program which orders the elements and
computes the function on every element by traversing the list only once. The
composition of these two programs is given by the following definition:
\begin{coqcode}
Definition qsort_times_two := Lmap times_two \o hylo merge qsplit.
\end{coqcode}
where \coq{Lmap times_two} is a list map function defined as a hylomorphism,
and \coq{times_two} multiplies every element of the list by two.
We can use Coq's generalised rewriting and \coq{hylo_fusion_l} to fuse
\coq{times_two} into \coq{hylo merge qsplit} which gives the program
\coq{hylo (merge \o natural times_two) qsplit}.
In this definition, \coq{natural} defines a natural transformation by
applying \coq{times_two} to the shapes thus multiplying every pivot by two.
%% Our formalisation contains a proof that \coq{natural} is indeed a natural
%% transformation, which relies on the fact that it preserves the structure of the
%% shapes and, therefore, the validity of the positions.
The extracted OCaml code is a single recursive traversal:
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
\todo{Write something here, don't leave it empty}{}

@@

\subsection{Shortcut Deforestation}\label{sec:shortcut}
The final case study we consider is shortcut deforestation on lists.  Shortcut
deforestation can be expressed succinctly in terms of hylomorphisms and their
laws~\cite{TakanoM95}. In particular, given a function:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
We can conclude, by parametricity, that
\begin{coqcode}
  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
\end{coqcode}
This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
\textbf{not} provable in Coq if we want to remain axiom-free, since we would
need to add the necessary parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
However, we prove a specific version of this theorem for the container whose
least fixed point is a list, and use it to encode the example from Takano and
Meijer~\cite{TakanoM95}:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
\end{coqcode}
Here, we are defining function \coq{sf1} as the composition of \coq{length},
\coq{Lmap f} and \coq{append ys}. Functions \coq{length} and \coq{Lmap} are
catamorphisms. Function \coq{append ys} is also a catamorphism that
appends \coq{ys} to an input list. It is defined by applying an algebra to
every cons node of a list, and applying a catamorphism with the input algebra
to \coq{ys} in the nil case:
\begin{coqcode}
Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
  fun x => match x with | MkCont sx kx => match sx with
  | s_nil => fun _ => (hylo a ilist_coalg) l
  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
  end kx end.
Definition append (l : list A) := hylo (tau l l_in) ilist_coalg.
\end{coqcode}
Here, \coq{ilist_coalg} is a recursive coalgebra from Coq lists to the
\coq{ListF} container. We apply the hylo fusion law repeatedly, unfold
definitions, and simplify in our specification for \coq{sf1}:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
  rewrite hylo_map_fusion, <- acid_rain. simpl; reflexivity.
Defined.
\end{coqcode}
From this, we extract the following OCaml code:
\begin{ocamlcode}
let rec sf1 f ys =
  function | [] -> let rec f0 = function
                   | [] -> (Uint63.of_int (0))
                   | _ :: t -> add (Uint63.of_int (1)) (f0 t)
                   in f0 ys
           | _ :: t -> add (Uint63.of_int (1)) (sf1 f ys t)
\end{ocamlcode}

>>>>

\newmintinline[coq]{coq}{fontsize=\small, breaklines}
\newmintinline[ocaml]{OCaml}{fontsize=\small}
\newminted{coq}{fontsize=\small}
\newminted{ocaml}{fontsize=\small}

@@

The container encoding \coq{ITreeF} has two shapes, one for leaves and one for
nodes, and nodes have two positions, one for each sublist \coq{l r : X} in
\coq{i_node}.
\begin{coqcode}
Inductive Tshape L A := | Leaf | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
The container for \coq{ITreeF} is denoted by \coq{TreeF}. For convenience, we
define the wrappers \coq{a_leaf} and \coq{a_node} to construct values of the
extension of this container.
Now, the container for quicksort is given by \coq{TreeF int}.
In the definition below, we use \coq{posL} and \coq{posR} as wrappers for
\coq{Lbranch} and \coq{Rbranch} with validity proofs.
\begin{coqcode}
Definition qsplit_f (x : list int) : App (TreeF int) (list int) :=
  match x with
  | nil => a_leaf tt
  | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in a_node h l r
  end.

Definition merge_f (x : App (TreeF int) (list int)) : list int :=
  let (sx, kx) := x in
  match sx return (Container.Pos sx -> _) -> _ with
  | Leaf _ _ => fun _ => nil
  | Node _ h => fun k => List.app (k (posL h)) (h :: k (posR h))
  end kx.
\end{coqcode}
The proofs that these are proper morphisms (called \coq{merge} and \coq{qsplit})
are automatically discharged by our tactics, but we still have to prove that the
coalgebra \coq{qsplit} is recursive. In this specific case we have to prove that
the two sublists are smaller than the input list. In general, we are required to
prove that the input is related to the data inside the structure functor, that
is that the element of the carrier of the coalgebra is related to the content of
the functor by some well-founded relation. Here we give a specialised property, the
full general property can be found in the formalisation:
\begin{coqcode}
Definition respects_relation `{setoid A} (c : Coalg F B) B} (R : B -> B -> Prop)
:= forall x (p : Pos (shape (c x))), R (length (cont (c x) p)) (length x).
\end{coqcode}
The resulting code extracted from \coq{hylo merge qsplit} is similar to a
hand-written implementation:
\begin{ocamlcode}
let rec qsort = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
    let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
    app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
The correctness proof is done by proving the following statement:
\begin{coqcode}
forall (l : list int), Sorted (hylo merge qsplit l) /\ Perm l (hylo merge qsplit l).
\end{coqcode}
that is any list produced by quicksort is sorted and is some permutation of the original one.

@@

\subsubsection{Fusing a divide-and-conquer computation}
We now show how we can use program calculation techniques to fuse a traversal
with a  divide-and-conquer algorithm to obtain a new program that only performs
recursion once instead of twice. In particular, we can map a function to the
result of quicksort and obtain a new program which orders the elements and
computes the function on every element by traversing the list only once. The
composition of these two programs is given by the following definition:
\begin{coqcode}
Definition qsort_times_two := Lmap times_two \o hylo merge qsplit.
\end{coqcode}
where \coq{Lmap times_two} is a list map function defined as a hylomorphism,
and \coq{times_two} multiplies every element of the list by two.
We can use Coq's generalised rewriting and \coq{hylo_fusion_l} to fuse
\coq{times_two} into \coq{hylo merge qsplit} which gives the program
\coq{hylo (merge \o natural times_two) qsplit}.
In this definition, \coq{natural} defines a natural transformation by
applying \coq{times_two} to the shapes thus multiplying every pivot by two.
%% Our formalisation contains a proof that \coq{natural} is indeed a natural
%% transformation, which relies on the fact that it preserves the structure of the
%% shapes and, therefore, the validity of the positions.
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
The extracted OCaml code is a single recursive traversal. Note the similarity to
a hand-written implementation, and that it has been derived by fusing two
different recursive programs using regular Coq rewrite tactics.

@@

\subsection{Shortcut Deforestation}\label{sec:shortcut}
The final case study we consider is shortcut deforestation on lists.  Shortcut
deforestation can be expressed succinctly in terms of hylomorphisms and their
laws~\cite{TakanoM95}. In particular, given a function:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
We can conclude, by parametricity, that
\begin{coqcode}
  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
\end{coqcode}
This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
\textbf{not} provable in Coq without adding the parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
However, we prove a specific version of this theorem for the container whose
least fixed point is a list, and use it to encode the example from Takano and
Meijer~\cite{TakanoM95}:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
\end{coqcode}
Here, we are defining function \coq{sf1} as the composition of \coq{length},
\coq{Lmap f} and \coq{append ys}. Functions \coq{length} and \coq{Lmap} are
catamorphisms. Function \coq{append ys} is also a catamorphism that
appends \coq{ys} to an input list. It is defined by applying an algebra to
every cons node of a list, and applying a catamorphism with the input algebra
to \coq{ys} in the nil case:
\begin{coqcode}
Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
  fun x => match x with | MkCont sx kx => match sx with
  | s_nil => fun _ => (hylo a ilist_coalg) l
  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
  end kx end.
Definition append (l : list A) := hylo (tau l l_in) ilist_coalg.
\end{coqcode}
Here, \coq{ilist_coalg} is a recursive coalgebra from Coq lists to the
\coq{ListF} container. We apply the hylo fusion law repeatedly, unfold
definitions, and simplify in our specification for \coq{sf1}:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
  rewrite hylo_map_fusion, <- acid_rain. simpl; reflexivity.
Defined.
\end{coqcode}
From this, we extract the following OCaml code:
\begin{ocamlcode}
let rec sf1 f ys =
  function | [] -> let rec f0 = function
                   | [] -> (Uint63.of_int (0))
                   | _ :: t -> add (Uint63.of_int (1)) (f0 t)
                   in f0 ys
           | _ :: t -> add (Uint63.of_int (1)) (sf1 f ys t)
\end{ocamlcode}
